Analyzing X-ray images in Python with scikit-image

### 2016-08-21 skimage_tomo.tex

\begin{abstract} % abstract
    The exploration and processing of images is a vital aspect of the
    scientific workflow of different X-ray imaging modalities. Users
    require tools combining interactivity, versatility, and performance.

    \texttt{scikit-image} is an open-source image processing toolkit for
    the Python language, that supports a large variety of file formats
    and is compatible with 2-D and 3-D images. The toolkit exposes a
    simple programming interface, with thematic modules grouping
    functions according to their purpose, such as image restoration,
    segmentation, and measurements. \texttt{scikit-image} users benefit
    from the rich ecosystem of Python for science, with powerful
    visualization libraries or machine learning toolkits.
    \texttt{scikit-image} combines a gentle
    learning curve, versatile image processing capabilities, and the
    scalable performances required for the high-throughput analysis of
    X-ray imaging data.

@@

Several software applications and libraries are available to synchrotron
users to process their images. ImageJ \citep{Abramoff2004, Schneider2012}
and its distribution Fiji \citep{Schindelin2012} is a popular tool for
2-D and 3-D images, thanks to its intuitive menus and graphical tools,
and the wealth of plugins contributed by a vivid
community \citep{Schindelin2015}. For 3-D images, commercial tools such
as Avizo 3D software (TM) are appreciated for an intuitive graphical
pipeline and advanced 3D visualization. Alternatively, the use of a
programming language gives finer control and more possibilities, provided
classical processing algorithms can be called from libraries -- hence
limiting the programming task and the risk of bugs. Matlab (TM) and its
image processing toolbox are popular in the academic community of
computer vision and image processing. The Python language is widely used
in the scientific world and in synchrotron facilities. As a
general-purpose language, Python is used in synchrotrons to control
device servers \citep{pytango}, to access raw data of X-ray
detectors \citep{Knudsen2013}, to reconstruct tomography volumes from
radiographs \citep{Gursoy2014, Mirone2014}, and in data
processing packages for macromolecular cristallography \citep{Adams2010},
azimuthal integration of diffraction data \citep{Ashiotis2015}, or
fluorescence analysis \citep{pymca}.

@@

In this section, we provide a short overview of the typical use patterns of
\texttt{scikit-image}, illutrated by short snippets of code. Indeed, since
Python is a programming language, the user interacts with data objects and
images through code, which is either typed and executed in an interactive
interpreter, or written in text files (scripts) that are executed.

\begin{figure*}
    \centerline{\includegraphics[width=0.8\textwidth]{ecosystem_landscape}}
    \caption{\csentence{\texttt{scikit-image} and the Scientific Python
	ecosystem}. Images are opened from files as \texttt{NumPy}
	arrays. Functions of \texttt{scikit-image} transform image arrays
	into other arrays with the same dimensions, or into arrays of
	numbers corresponding to features of the image. The output of
	scikit-image functions can be passed to other Python modules
	relying on \texttt{NumPy} arrays, such as \texttt{SciPy} or
	\texttt{scikit-learn}. Image-shaped arrays are transformed into
	visualizations with \texttt{matplotlib} (2D) or \texttt{Mayavi}
	(3D). A variety of environments is available for code development
	and execution, from classical IDEs to Jupyter notebooks.
    \label{fig:ecosystem}}
\end{figure*}

Images are manipulated as numerical arrays of uniform
data type, a format that guarantees a fast access to the computer memory. The
n-dimensional (2-D, 3-D, \dots) numerical array object is provided by the
\texttt{NumPy} module \citep{Vanderwalt2011}.

In image processing, one of the first tasks is therefore to generate
\texttt{NumPy} arrays, which is often done by reading data from files. We
read one 2-D image from a file and display it with the following lines of code:

@@

\texttt{skimage} is the name under which \texttt{scikit-image} is imported in
Python code. Note that functions (such as \texttt{imread} to read an image
file, or \texttt{imshow} to display an image) are found in thematic submodules of
\texttt{skimage}, such as \texttt{io} for Input/Output.

A stack of 2-D images, such as tomography slices generated by a reconstruction algorithm, can be opened as an image collection or a 3-D array:
\lstinputlisting[language=Python]{io_coll.py}
Raw data formats can be opened using the \texttt{NumPy} functions \texttt{fromfile} or \texttt{memmap}, and \texttt{hdf5} files are opened using modules such as \texttt{h5py} or \texttt{pytables}.

\texttt{scikit-image} has a simple Application Programming Interface
(API), based almost only on
functions. Most functions take an image (\emph{i.e} a multi-dimensional
array) as input parameter:
\begin{lstlisting}
>>> from skimage import filters
>>> im_gaussian = filters.gaussian(im)
\end{lstlisting}

Optional parameters can be passed as Python's \emph{keyword arguments},
in addition to the image parameter.
\begin{lstlisting}
>>> im_gaussian = filters.gaussian(im, sigma=3, mode='wrap')
\end{lstlisting}
A few functions require several arrays to be passed, such as the
watershed segmentation algorithm that takes as parameters the image to be
segmented, and an image of markers from which labels are propagated:
\begin{lstlisting}
>>> from skimage import morphology
>>> labels = morphology.watershed(im, markers)
\end{lstlisting}
Therefore, the image processing workflow can be considered as a directed graph
(a richer structure than a linear pipeline), where nodes are image-shaped arrays, and
edges are functions of \texttt{scikit-image} transforming the arrays (see
Fig.~\ref{fig:ecosystem}).

Most functions transparently handle 2-D, 3-D, or even higher-dimensional
images as arguments, so
the same functions can be used to process tomography, microscopy, or natural
images. A few functions do not support 3-D images and raise an error
when passed a 3-D argument:
\begin{lstlisting}
>>> filters.prewitt(im)
[...]
ValueError: The parameter `image` must be a 2-dimensional array
\end{lstlisting}

However, the proportion of functions supporting 3D images is always increasing,
thanks to the many contributors to the library.

A subset of functions returns not images but other value types,
such as pixel coordinates of objects of interest, or statistical
information about the image:
\begin{lstlisting}
>>> from skimage import exposure
>>> counts, bins = exposure.histogram(im)
>>> counts.shape
(256,)
\end{lstlisting}

@@

While several interpreters are available to execute Python
instructions and scripts, the most popular one in the scientific world is
the IPython interpreter \citep{Perez2007, Rossant2015}. IPython is an
advanced interpreter, which integrates syntax highlighting, text auto-completion,
a debugger,
introspection and profiling methods, and online help. Several
Integrated Development Environments (IDEs) come bundled with IPython,
together with other components such as a text editor. Notable examples
include Spyder (Fig.~\ref{fig:spyder}), PyCharm, and Visual Studio Code.

@@

The Jupyter notebook \citep{Kluyver2016} is a web application that grew
from the IPython project. Jupyter notebooks provide a development
environment within a web browser, where live code can be enriched
by explanatory text, equations and visualizations (FIGURE).
As of July 2016, more than 500,000 Jupyter notebooks were posted on GitHub,
demonstrating their wide adoption by the community as workflow-sharing
tools (http://archive.ipython.org/media/SciPy2016JupyterLab.pdf).

\paragraph{NumPy arrays} are the cornerstone of the Scientific Python
ecosystem, and of \texttt{scikit-image} operations in particular.
Cropping or downsampling an image, or retrieving pixels corresponding to
a given label in a segmentation are all one-liners NumPy code. To
illustrate the compactness of NumPy code, modifying pixels values below a
threshold can be written as
\begin{lstlisting}
im[im < 0.5] = 0
\end{lstlisting}
using the ability to index arrays with boolean arrays, called
\emph{masking}. \texttt{NumPy} uses memory sparingly and avoids
making new copies of arrays whenever possible, an important
requirement when dealing with the gigabyte-sized images of tomography.
For example, cropping a subvolume as follows does not create a copy of the
original array:
\begin{lstlisting}
sub_volume = im[100:-100, 100:-100, 100:-100]
\end{lstlisting}

@@

An easy solution consists in dividing an image into smaller images (with
or without overlap, depending on the operation), and to apply the
same operation on the different sub-images, on different cores. Creating overlapping chunks is easy with the dedicated function \texttt{view\_as\_windows} (or \texttt{view\_as\_blocks} for contiguous non-overlapping chunks):
\begin{lstlisting}[language=Python]
>>> from skimage import util, data
>>> im = data.camera()
>>> im.shape
(512, 512)
>>> # chunks of size 120 with 8-pixel overlap
>>> chunks = util.view_as_windows(im, (128, 128), step=120)
>>> chunks.shape
(4, 4, 128, 128)
\end{lstlisting}
The \texttt{joblib} library enables easy parallel processing. Looping over the different blocks, and dispatching the computation over several cores, is realized with the following syntax:
\begin{lstlisting}
>>> from joblib import Parallel, delayed
>>> filtered_chunks = Parallel(n_jobs=4)(delayed(filters.gaussian)(chunks[i, j]) for i in range(4) for j in range(4))
\end{lstlisting}
\texttt{scikit-image} also offers experimental support for a more integrated parallel processing pipeline, thanks to the \texttt{dask} \citep{dask} module:
\begin{lstlisting}
filtered_im = util.apply_parallel(filters.gaussian, im, depth=8)
\end{lstlisting}
The size of chunks is determined automatically from the number of available cpus, or can be specified by the user.

>>>>

\begin{abstract} % abstract
    The exploration and processing of images is a vital aspect of the
    scientific workflows of many X-ray imaging modalities. Users
    require tools that combine interactivity, versatility, and performance.

    \texttt{scikit-image} is an open-source image processing toolkit for
    the Python language that supports a large variety of file formats
    and is compatible with 2-D and 3-D images. The toolkit exposes a
    simple programming interface for functions that perform various
    operations on images, grouped thematically according to their
    purpose, e.g. image restoration,
    segmentation, measurements, etc. \texttt{scikit-image} users benefit
    from a rich scientific Python ecosystem that contains many
    powerful libraries for tasks such as visualization or machine learning.
    \texttt{scikit-image} provides a gentle
    learning curve, versatile image processing capabilities, and the
    scalable performances required for the high-throughput analysis of
    X-ray imaging data.

@@

Several software applications and libraries are available to synchrotron
users to process their images. ImageJ \citep{Abramoff2004, Schneider2012}
and its distribution Fiji \citep{Schindelin2012} is a popular tool for
2-D and 3-D images, thanks to its intuitive menus and graphical tools,
and the wealth of plugins contributed by a vivid
community \citep{Schindelin2015}. For 3-D images, commercial tools such
as Avizo 3D software (TM) are appreciated for an intuitive graphical
pipeline and advanced 3D visualization. Alternatively, the use of a
programming language gives finer control and more possibilities, provided
classical processing algorithms can be called from libraries -- thereby
limiting the complexity of the programming task and the risk of bugs. Matlab (TM) and its
image processing toolbox are popular in the academic community of
computer vision and image processing. The Python language is widely used
in the scientific world and in synchrotron facilities. As a
general-purpose language, Python is used in synchrotrons to control
device servers \citep{pytango}, to access raw data of X-ray
detectors \citep{Knudsen2013}, to reconstruct tomography volumes from
radiographs \citep{Gursoy2014, Mirone2014}, and in data
processing packages for macromolecular cristallography \citep{Adams2010},
azimuthal integration of diffraction data \citep{Ashiotis2015} or
fluorescence analysis \citep{pymca}.

@@

In this section, we provide a short overview of the typical use patterns of
\texttt{scikit-image}, illutrated by short snippets of code. Since
Python is a programming language, the user interacts with data objects and
images through code, which is either entered and executed in an interactive
interpreter, or written in text files (so-called scripts) that are executed.

\begin{figure*}
    \centerline{\includegraphics[width=0.8\textwidth]{ecosystem_landscape}}
    \caption{\csentence{\texttt{scikit-image} and the Scientific Python
	ecosystem}. Images are opened from files as \texttt{NumPy}
	arrays. Functions of \texttt{scikit-image} transform image arrays
	into other arrays with the same dimensions, or into arrays of
	numbers corresponding to features of the image. The output of
	scikit-image functions can be passed to other Python modules
	relying on \texttt{NumPy} arrays, such as \texttt{Scipy} or
	\texttt{scikit-learn}. Image-shaped arrays are transformed into
	visualizations with \texttt{matplotlib} (2D) or \texttt{Mayavi}
	(3D). A variety of environments is available for code development
	and execution, from classical IDEs to Jupyter notebooks.
    \label{fig:ecosystem}}
\end{figure*}

Images are manipulated as numerical arrays, each with a single, uniform
data type.  This common format guarantees interoperability with other
libraries and straightforward access to and interpretation of computer memory. The
N-dimensional (2-D, 3-D, \dots) numerical array object is provided by the
\texttt{NumPy} module \citep{Vanderwalt2011}.

In image processing in Python, one of the first tasks then is to generate
\texttt{NumPy} arrays, which is often achieved by reading data from files. We
read one 2-dimensional image from a file and display it as follows:

@@

\texttt{skimage} is the name under which \texttt{scikit-image} is imported in
Python code. Note that functions (such as \texttt{imread} that reads an image
file, or \texttt{imshow} that displays an image) are found in thematic submodules of
\texttt{skimage}, such as \texttt{io} for Input/Output.

A stack of 2-D images, such as tomography slices generated by a reconstruction algorithm, can be opened as an image collection or a 3-D array:
\lstinputlisting[language=Python]{io_coll.py}
Raw data formats can be opened using the \texttt{NumPy} functions \texttt{fromfile} or \texttt{memmap}, and \texttt{hdf5} files are accessed using modules such as \texttt{h5py}, \texttt{pytables}.

\texttt{scikit-image} has a very simple Application Programming Interface
(API), based almost exclusively on
functions. Most functions take an image (\emph{i.e} a multi-dimensional
array) as input parameter:
\begin{lstlisting}
>>> from skimage import filters
>>> im_gaussian = filters.gaussian(im)
\end{lstlisting}

Optional parameters can be passed as Python \emph{keyword arguments},
in addition to the image parameter.
\begin{lstlisting}
>>> im_gaussian = filters.gaussian(im, sigma=3, mode='wrap')
\end{lstlisting}
A few functions require several arrays to be passed, such as the
watershed segmentation algorithm that takes as parameters the image to be
segmented, and an image of markers from which labels are propagated:
\begin{lstlisting}
>>> from skimage import morphology
>>> labels = morphology.watershed(im, markers)
\end{lstlisting}
Therefore, the image processing workflow can be seen as a directed graph
(a richer structure than a linear pipeline), where nodes are image-shaped arrays, and
edges are functions of \texttt{scikit-image} transforming the arrays (see
Fig.~\ref{fig:ecosystem}).

Most functions take both 2-D and 3-D images as arguments, so that
the same functions can be used to process tomography and microscopy
images. The rest raise an error
when passed a 3-D argument:
\begin{lstlisting}
>>> filters.prewitt(im)
[...]
ValueError: The parameter `image` must be a 2-dimensional array
\end{lstlisting}

While the majority of functions return processed images, returns can
also be numerical value(s) such as pixel coordinates of objects of
interest or statistical
information about the image:
\begin{lstlisting}
>>> from skimage import exposure
>>> counts, bins = exposure.histogram(im)
>>> counts.shape
(256,)
\end{lstlisting}

@@

While several interpreters are available to execute interactively Python
instructions and scripts, the most popular one in the scientific world is
the IPython interpreter \citep{Perez2007, Rossant2015}. IPython is an
advanced interpreter, which integrates syntax highlighting, text auto-completion,
a debugger,
introspection and profiling methods, and online help. Several
Integrated Development Environments (IDEs) include IPython,
together with other components such as a text editor. Notable examples
include Spyder (Fig.~\ref{fig:spyder}), PyCharm, and Visual Studio Code.

@@

The Jupyter notebook \citep{Kluyver2016} is a web application that grew
from the IPython project. Jupyter notebooks provide an interactive development
environment within a web browser, where live code can be enriched
by explanatory text, equations and visualizations (FIGURE).
(jni: FURTHER TEXT ABOUT IMPACT OF JUPYTER NOTEBOOKS)

\paragraph{NumPy arrays} are the cornerstone of the Scientific Python
ecosystem, and of \texttt{scikit-image} operations in particular.
Cropping or downsampling an image, or retrieving pixels corresponding to
a given label in a segmentation are all one-liners NumPy code. To
illustrate the compactness of NumPy code, modifying pixels values below a
threshold can be written as
\begin{lstlisting}
im[im < 0.5] = 0
\end{lstlisting}
using the possibility to index arrays with boolean arrays, called
\emph{masking}. Also, \texttt{NumPy} uses memory sparingly and avoids
making new copies of arrays whenever possible, which is a important
requirement when dealing with the gigabyte-sized images of tomography.
For example, cropping a subvolume as follows does not create a copy of the
original array:
\begin{lstlisting}
sub_volume = im[100:-100, 100:-100, 100:-100]
\end{lstlisting}

@@

An easy solution consists in dividing an image into smaller images (with
or without overlap, depending on the operation), and to apply the
same operation on the different sub-images, on different cores. Creating overlapping chunks is easy with the dedicated function \texttt{view\_as\_windows} (or \texttt{view\_as\_blocks} for contiguous non-overlapping chunks):
\begin{lstlisting}[language=Python]
>>> from skimage import util, data
>>> im = data.camera()
>>> im.shape
(512, 512)
>>> # chunks of size 120 with 8-pixel overlap
>>> chunks = util.view_as_windows(im, (128, 128), step=120)
>>> chunks.shape
(4, 4, 128, 128)
\end{lstlisting}
The \texttt{joblib} module enables easy parallel processing. Looping over the different blocks, and dispatching the computation over several cores, is realized with the following syntax:  
\begin{lstlisting}
>>> from joblib import Parallel, delayed
>>> filtered_chunks = Parallel(n_jobs=4)(delayed(filters.gaussian)(chunks[i, j]) for i in range(4) for j in range(4))
\end{lstlisting}
\texttt{scikit-image} also offers experimental support for a more integrated parallel processing pipeline, thanks to the \texttt{dask} \citep{dask} module:
\begin{lstlisting}
filtered_im = util.apply_parallel(filters.gaussian, im, depth=8)
\end{lstlisting}
The size of chunks is determined automatically from the number of available cpus, or can be specified by the user.
